---
title: "Exploring OD2017 data"
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)

library(tidyverse)
library(piggyback)
library(sf)
library(tmap)
library(leaflet)
library(stplanr)
library(od)
library(osrm)
# assume data is in the folder
# folder = "/home/lucas/"
# folder = "~/Dropbox/DATA/"
folder = "~/hd/data/project-data/spshared/"

pb_download(repo = "spstreets/OD2017",
            tag = "1",
            dest = ".")

# folder = "~/hd/data/project-data/spshared/"
# load(file.path(folder, "OD2017.rda"))
OD_2017_v1 = readRDS("OD_2017_v1.Rds")

```

I create this separated document to be the main reference on the exploratory data analysis. The data set is at the trip level and I take advantage of the survey design and use the expansion factor (for the trips) in all the graphs and analysis here. The survey implies that there are `r sum(OD_2017_v1$fe_via, na.rm=TRUE)` trips/day in the São Paulo Metro Region.

### Key variables

The data set contains a lot of socioeconomic variables. Since our interest relies on trips, flows and mode choices, the analyses so far are at the trip level. In this context, the variables that contain a the modes used in each trip are the main ones (`modo1`, `modo2`, `modo3`, and `modo4`). There is also the coordinates for each residence (`co_dom_x` and `co_dom_y`) and workplaces (`co_tr1_x` and `co_tr1_y`), and obviously the origin and destination zones and coordinates (e.g. `zona_o`, `co_o_x`, `co_o_y`) and euclidean distance. As you can see in the graph below, the vast majority of trips are made using just one mode. I use the main mode used in the trip (`modoprin`) in the remaining graphs. 

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  mutate(
    n_modes = case_when(
      !is.na(modo1) & is.na(modo2) & is.na(modo3) & is.na(modo4)    ~ "1",
      !is.na(modo1) & !is.na(modo2) & is.na(modo3) & is.na(modo4)   ~ "2",
      !is.na(modo1) & !is.na(modo2) & !is.na(modo3) & is.na(modo4)  ~ "3",
      !is.na(modo1) & !is.na(modo2) & !is.na(modo3) & !is.na(modo4) ~ "4"
    )
  )
  

OD_2017_v1 %>%
  ggplot() +
  geom_bar(aes(n_modes, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Number of modes used in each trip") +
  xlab("") +
  theme_bw()

```


### Different ways of defining modes

I aggregate the different names used for similar things. And this is the resulting distribution of trips for each mode.

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    modo_simples = case_when(
      modoprin==1 | modoprin==2 | modoprin==3 ~ "Metrô, trem e monotrilho",
      modoprin==4 | modoprin==5 | modoprin==6 ~ "Ônibus",
      modoprin==7 | modoprin==8            ~ "Fretado e Escolar",
      modoprin==9 | modoprin==10           ~ "Automóvel",
      modoprin==11 | modoprin==12          ~ "Táxi e derivados",
      modoprin==13 | modoprin==14          ~ "Moto",
      modoprin==15                      ~ "Bicicleta",
      modoprin==16                      ~ "A pé",
      modoprin==17                      ~ "Outros",
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(modo_simples, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Aggregating similar modes") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```


### Recategorisation of modes

I tried two different recategorisations (see code for the choices).The first is based on the "vehicle" itself, and the other relies on the "role" of the individual while in the trip.

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    mode_recat1 = case_when(
      modoprin %in% 1:6 ~ "Public transport",
      modoprin %in% 9:10 ~ "Car",
      modoprin %in% 13:14 ~ "Motorcycle",
      modoprin == 15 ~ "Bicycle",
      modoprin == 16 ~ "On foot",
      modoprin == 17 | modoprin == 12 | modoprin == 11 | modoprin == 7 | modoprin == 8 ~ "Other"
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(mode_recat1, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Recategorisation of modes 1") +
  xlab("")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    mode_recat2 = case_when(
      modoprin %in% 1:6 ~ "Public transport",
      modoprin == 9 | modoprin == 13 ~ "Driving",
      modoprin==10 | modoprin==14 ~ "Rider",
      modoprin %in% c(7, 8, 11, 12) ~ "Private transport passenger",
      modoprin==15 ~ "Cycling",
      modoprin==16 ~ "Walking",
      modoprin==17 ~ "Other"
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(mode_recat2, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Recategorisation of modes - 2") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```

Now the A/B Streets classification (foot, bike, car, public transport and other):

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    mode_ab_streets = case_when(
      modoprin %in% 1:6 ~ "Public transport",
      modoprin == 16 ~ "Foot",
      modoprin==9 | modoprin==10 ~ "Car",
      modoprin %in% c(7, 8, 11, 12, 13, 14, 17) ~ "Other",
      modoprin==15 ~ "Bike"
    )
  )


OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(mode_ab_streets, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("A/B Streets Classification") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))


```

Now the mode share by distance (using the first recategorisation). The average trips' distance is `r weighted.mean(OD_2017_v1$distancia, OD_2017_v1$fe_via, na.rm = TRUE)`. 


```{r, echo=FALSE}

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  mutate(
    dist_km = distancia/1000,
    dist_bands = cut(dist_km,
                     breaks = c(0, 1, 3, 6, 10, 15, 20, 30, 83),     # max is 82.7
                     include.lowest = TRUE)
  ) %>%
  select(dist_bands, mode_recat1, fe_via) %>%
  group_by(dist_bands, mode_recat1) %>%
  summarise(Trips = sum(fe_via)) %>%
  ggplot(aes(x = dist_bands, y = Trips, fill = mode_recat1)) +
  geom_col() +
  xlab("Distance (km)") +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)) +
  scale_fill_brewer(palette = "Set1", name = "Mode") +
  theme_bw()

```


### [OFF] Checking whether the coordinates are real or just sub-zones centroids.


```{r}

n_dom_coords = OD_2017_v1 %>%
  group_by(co_dom_x, co_dom_y) %>%
  count()

```

There are `r length(unique(OD_2017_v1$id_dom)) - dim(n_dom_coords)[1]` more households than households' coordinates pairs.


### Basic temporal analysis

Distributions of trips in the day.

```{r, echo=FALSE, fig.cap = "Using `motivo_o`, reason at the origin, to define."}

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  mutate(
    saida = h_saida + min_saida/60,
    chegada = h_cheg + min_cheg/60,
    motivo = case_when(
      motivo_o %in% 1:3 ~ "Trabalho",
      motivo_o == 4 ~ "Escola",
      motivo_o == 8 ~ "Residência",
      TRUE ~ "Outro"
    )
  ) %>%
  ggplot() +
  geom_density(aes(x=saida, weight=fe_via, linetype="Saída")) +
  geom_density(aes(x=chegada, weight=fe_via, linetype="Chegada")) +
  scale_linetype_manual(name="",
                        values = c("Saída"="solid", "Chegada"="dashed")) +
  geom_vline(xintercept = c(10, 15), linetype = "dashed", colour = "red", size = 1) +
  ylab("Trips (density)") +
  xlab("Hour") +
  facet_wrap(~motivo) +
  theme_bw()

```


```{r, echo=FALSE, fig.cap = "Using `motivo_d`, reason at the destination, to define."}

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  mutate(
    saida = h_saida + min_saida/60,
    chegada = h_cheg + min_cheg/60,
    motivo = case_when(
      motivo_d %in% 1:3 ~ "Trabalho",
      motivo_d == 4 ~ "Escola",
      motivo_d == 8 ~ "Residência",
      TRUE ~ "Outro"
    )
  ) %>%
  ggplot() +
  geom_density(aes(x=saida, weight=fe_via, linetype="Saída")) +
  geom_density(aes(x=chegada, weight=fe_via, linetype="Chegada")) +
  scale_linetype_manual(name="",
                        values = c("Saída"="solid", "Chegada"="dashed")) +
  geom_vline(xintercept = c(10, 15), linetype = "dashed", colour = "red", size = 1) +
  ylab("Trips (density)") +
  xlab("Hour") +
  facet_wrap(~motivo) +
  theme_bw()

```


### Basic geographic analysis (now with times  )

```{r, include=FALSE}
# # list.files(folder)
# zonas_od = st_read("~/hd/data/project-data/spshared//shapefiles/Zonas_2017_region.shp")
# zonas_od = st_read(file.path(folder, "shapefiles/Zonas_OD2017/Zonas_2017_region.shp")
#                    , options = "ENCODING=LATIN1"
#                    )
# sf::st_write(zonas_od, "zonas_od.gpkg")
# piggyback::pb_upload("zonas_od.gpkg")
# zonas_od = sf::read_sf("zonas_od.gpkg")
# piggyback::pb_download("zonas_od.Rds")
zonas_od = readRDS("zonas_od.Rds")
tmap_mode("plot")         # mode = view breaks some plots (why? ask Robin)

```

Here I plot the main zones of origin and destination using the total number of trips -- always using the survey's expansion factor. It is interesting to note that there is no centrality pattern when we look at the main Zones of Destination. Maybe we should restrict the analysis to specific types of trips (e.g., commute strictly defined). I checked these results with the official Tables and they are the same in terms of the main zones of origin and destination.

```{r, echo=FALSE}
OD = OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  select(zona_o, zona_d, modoprin, h_saida, fe_via) %>%
  mutate(
    public = ifelse(modoprin %in% 1:6, 1, 0)*fe_via,
    car = ifelse(modoprin %in% 9:10, 1, 0)*fe_via,
    motorcycle = ifelse(modoprin %in% 13:14, 1, 0)*fe_via,
    bike = ifelse(modoprin == 15, 1, 0)*fe_via,
    foot = ifelse(modoprin == 16, 1, 0)*fe_via,
    other = ifelse(modoprin %in% c(7, 8, 11, 12, 17), 1, 0)*fe_via,
    time = case_when(
      h_saida >= 0 & h_saida < 10 ~ "Morning",
      h_saida >= 10 & h_saida < 15 ~ "Noon",
      h_saida >=15 & h_saida <= 23 ~ "Evening"
    )
         ) %>%
  select(zona_o, zona_d, time, public, car, motorcycle, bike, foot, other) %>%
  group_by(zona_o, zona_d, time) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(trips = public + car + motorcycle + bike + foot + other)

# check:
# sum(OD$public) + sum(OD$car) + sum(OD$motorcycle) + sum(OD$bike) + sum(OD$foot) + sum(OD$other)

origin = OD %>%
  group_by(zona_o, time) %>%
  summarise_all(sum) %>%
  select(zona_o, time, trips_o = trips) %>%
  ungroup() %>%
  left_join(zonas_od, by = c("zona_o" = "NumeroZona")) %>%
  st_as_sf(sf_column_name = "geometry") %>%
  mutate(area = units::set_units(st_area(.), km^2),
         trips_o_km2 = as.numeric(trips_o/area)
         )

destination = OD %>%
  group_by(zona_d, time) %>%
  summarise_all(sum) %>%
  select(zona_d, time, trips_d = trips) %>%
  ungroup() %>%
  left_join(zonas_od, by = c("zona_d" = "NumeroZona")) %>%
  st_as_sf(sf_column_name = "geometry") %>%
  mutate(area = units::set_units(st_area(.), km^2),
         trips_d_km2 = as.numeric(trips_d/area)
         )
  
```

Plotting the origin zones by time.

```{r, echo=FALSE}

tm_shape(origin) + tm_fill("trips_o_km2", 
                           style = "cont",
                           title = "Trips/km²")  +
  tm_facets(by="time", nrow=3) +
  tm_layout(title = "São Paulo Metro Region - Origins",
            inner.margins = c(0, .02, .02, .02),
            legend.outside = TRUE)

```

Plotting the destination zones by time.

```{r, echo=FALSE}

tm_shape(destination) + tm_fill("trips_d_km2", 
                                style = "cont",
                                title = "Trips/km²") +
  tm_facets(by="time", nrow=3) +
  tm_layout(title = "São Paulo Metro Region - Destinations",
            inner.margins = c(0, .02, .02, .02),
            legend.outside = TRUE)

```

Looking at the São Paulo City. First, the origins.

```{r, echo=FALSE}

origin %>%
  filter(NomeMunici == "São Paulo") %>%
  tm_shape() +
  tm_fill("trips_o_km2",
          style = "cont",
          title = "Trips/km²") +
  tm_facets(by="time", ncol=3) +
  tm_layout(title = "São Paulo City - Origins",
            inner.margins = c(0, .02, .02, .02),
            legend.outside = TRUE)

```

Now, the destinations for São Paulo City.

```{r, echo=FALSE}

destination %>%
  filter(NomeMunici == "São Paulo") %>%
  tm_shape() +
  tm_fill("trips_d_km2",
          style = "cont",
          title = "Trips/km²") +
  tm_facets(by="time", ncol=3) +
  tm_layout(title = "São Paulo City - Destinations",
            inner.margins = c(0, .02, .02, .02),
            legend.outside = TRUE)

```


#### Active travels and trip patterns

(I removed the previous visualizations since they were not saying anything about the data.)

Walking and cycling are highly concentrated in few OD pairs (>90% in 1,000 pairs for walking and >90% in 500 pairs for cycling). Car trips are more dispersed, we have >90% of trips in 10,000 pairs. 

```{r, include=FALSE}

bike_pairs = OD %>%
  select(-time) %>%
  group_by(zona_o, zona_d) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(bike_share = (bike/trips)*100) %>%
  slice_max(bike,
            n = 500,
            with_ties = FALSE)

foot_pairs = OD %>%
  select(-time) %>%
  group_by(zona_o, zona_d) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(foot_share = (foot/trips)*100) %>%
  slice_max(foot,
            n = 1000,
            with_ties = FALSE)

car_pairs = OD %>%
  select(-time) %>%
  group_by(zona_o, zona_d) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(car_share = (car/trips)*100) %>%
  slice_max(car,
            n = 10000,
            with_ties = FALSE)

```


Bike trips:

```{r, echo=FALSE}

desire_lines_bike = od2line(bike_pairs, zonas_od)

tm_shape(desire_lines_bike) +
  tm_lines(palette = "plasma",
           breaks = c(0, 10, 25, 50, 75, 100),
           lwd = "trips",
           scale = 15,
           title.lwd = "Number of trips",
           alpha = 0.8,
           col = "bike_share",
           title = "Bike trips (%)") +
  tm_shape(zonas_od) +
  tm_borders(col = "lightgrey", lwd = 0.2) +
  tm_scale_bar()

```

Foot trips:

```{r, echo=FALSE}

desire_lines_foot = od2line(foot_pairs, zonas_od)

tm_shape(desire_lines_foot) +
  tm_lines(palette = "plasma",
           breaks = c(0, 10, 25, 50, 75, 100),
           lwd = "trips",
           scale = 15,
           title.lwd = "Number of trips",
           alpha = 0.8,
           col = "foot_share",
           title = "Foot trips (%)") +
  tm_shape(zonas_od) +
  tm_borders(col = "lightgrey", lwd = 0.2) +
  tm_scale_bar()

```

Car trips, still too many lines (10,000) to account for >92% of total car trips.

```{r, echo=FALSE}

desire_lines_car = od2line(car_pairs, zonas_od)

tm_shape(desire_lines_car) +
  tm_lines(palette = "plasma",
           breaks = c(0, 10, 25, 50, 75, 100),
           lwd = "trips",
           scale = 15,
           title.lwd = "Number of trips",
           alpha = 0.8,
           col = "car_share",
           title = "Car trips (%)") +
  tm_shape(zonas_od) +
  tm_borders(col = "lightgrey", lwd = 0.2) +
  tm_scale_bar()

```

Using the first 5,000 OD pairs, we have >78% of all car trips.

```{r, echo=FALSE}

car_pairs_5 = desire_lines_car %>%
    slice_max(car,
              n = 5000,
              with_ties = FALSE)


desire_lines_car = od2line(car_pairs_5, zonas_od)

tm_shape(desire_lines_car) +
  tm_lines(palette = "plasma",
           breaks = c(0, 10, 25, 50, 75, 100),
           lwd = "trips",
           scale = 15,
           title.lwd = "Number of trips",
           alpha = 0.8,
           col = "car_share",
           title = "Car trips (%)") +
  tm_shape(zonas_od) +
  tm_borders(col = "lightgrey", lwd = 0.2) +
  tm_scale_bar()


```

Using the biggest 1,000 pairs, we have >45% of all car trips.

```{r, echo=FALSE}

car_pairs_1 = desire_lines_car %>%
    slice_max(car,
              n = 1000,
              with_ties = FALSE)


desire_lines_car = od2line(car_pairs_1, zonas_od)

tm_shape(desire_lines_car) +
  tm_lines(palette = "plasma",
           breaks = c(0, 10, 25, 50, 75, 100),
           lwd = "trips",
           scale = 15,
           title.lwd = "Number of trips",
           alpha = 0.8,
           col = "car_share",
           title = "Car trips (%)") +
  tm_shape(zonas_od) +
  tm_borders(col = "lightgrey", lwd = 0.2) +
  tm_scale_bar()


```


### Main routes by mode (takes a long time here) -- rewriting this part.

```{r, eval=FALSE}

bike_routes = od_jitter(bike_pairs, zonas_od) %>%
  st_transform(crs = 4326) %>%
  mutate(route = st_geometry(line2route(.,
                                        route_fun = route_osrm,
                                        osrm.profile = "bike",
                                        time_delay = 1)
                             )
         )

st_geometry(bike_routes) = bike_routes$route

saveRDS(bike_routes, file = "bike_routes_2021-09-16.Rds")

```

```{r, eval=FALSE}

foot_routes = od_jitter(foot_pairs, zonas_od) %>%
  st_transform(crs = 4326) %>%
  mutate(route = st_geometry(line2route(.,
                                        route_fun = route_osrm,
                                        osrm.profile = "foot",
                                        time_delay = 0.5)
                             )
         )

st_geometry(foot_routes) = foot_routes$route

saveRDS(foot_routes, file = "foot_routes_2021-09-16.Rds")

```

```{r, eval=FALSE}

car_routes = od_jitter(car_pairs, zonas_od) %>%
  st_transform(crs = 4326) %>%
  mutate(route = st_geometry(line2route(.,
                                        route_fun = route_osrm,
                                        osrm.profile = "car",
                                        time_delay = 0.02)
                             )
         )

st_geometry(car_routes) = car_routes$route

saveRDS(car_routes, file = "car_routes_2021-09-16.Rds")

```



### Plotting the routes

Load the routes I calculated.

```{r, echo=FALSE}
piggyback::pb_download("bike_routes_2021-09-16.Rds")
bike_routes = readRDS("bike_routes_2021-09-16.Rds")
foot_routes = readRDS("foot_routes_2021-09-16.Rds")
car_routes = readRDS("car_routes_2021-09-16.Rds")           
```

something wrong here, check later...

```{r, eval=FALSE}
# remotes::install_cran("archive")
# archive::archive("SAD69-96_SHP_redecicloviaria.rar") # fails
# unrar SAD69-96_SHP_redecicloviaria.rar = bash command
# bike_paths = st_read("SAD69-96_SHP_redecicloviaria.shp")
# st_crs(bike_paths) = 5533     
# bike_paths = st_transform(bike_paths, crs = 4326)
# sf::write_sf(bike_paths, "bike_paths.gpkg")
# piggyback::pb_upload("bike_paths.gpkg")
bike_paths = read_sf("bike_paths.gpkg")
```


```{r, eval=FALSE}
route_net_bike = overline(sl = bike_routes, attrib = "bike")

leaflet(route_net_bike) %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
#  addPolylines(data = bike_paths, color = "green", weight = 1) %>%
  addPolylines(weight = route_net_bike$bike / 1000, color = "white")

```

```{r, eval=FALSE}

route_net_car = overline(sl = car_routes, regionalise = 9e99, attrib = "car")

leaflet(route_net_car) %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addPolylines(weight = route_net_car$car / 10000, color = "white")

```

```{r, eval=FALSE}

route_net_foot = overline(sl = foot_routes, regionalise = 9e99, attrib = "foot")

leaflet(route_net_foot) %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addPolylines(weight = route_net_foot$foot / 10000, color = "white")

```


### Infrastructure data (São Paulo municipality only)

```{r, echo=FALSE}

OD_sp = OD_2017_v1 %>%
  filter(!is.na(modoprin) &
           (muni_d == 36 & muni_o == 36)
         ) %>%
  select(zona_o, zona_d, modoprin, fe_via) %>%
  mutate(
    public = ifelse(modoprin %in% 1:6, 1, 0)*fe_via,
    car = ifelse(modoprin %in% 9:10, 1, 0)*fe_via,
    motorcycle = ifelse(modoprin %in% 13:14, 1, 0)*fe_via,
    bike = ifelse(modoprin == 15, 1, 0)*fe_via,
    foot = ifelse(modoprin == 16, 1, 0)*fe_via,
    other = ifelse(modoprin %in% c(7, 8, 11, 12, 17), 1, 0)*fe_via
    ) %>%
  select(zona_o, zona_d, public, car, motorcycle, bike, foot, other) %>%
  group_by(zona_o, zona_d) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(trips = public + car + motorcycle + bike + foot + other)

```

```{r, echo=FALSE}
# I have two identical geometries in each route dataframe
car_routes$route = NULL
bike_routes$route = NULL
foot_routes$route = NULL

bike_routes$tot_length = (units::set_units(st_length(bike_routes), km))*bike_routes$bike
car_routes$tot_length = (units::set_units(st_length(car_routes), km))*car_routes$car
foot_routes$tot_length = (units::set_units(st_length(foot_routes), km))*foot_routes$foot

```

obs.: very strange behavior here: st_buffer + st_make_valid makes one single polygon very small. without st_buffer, R does not even plot -- "C stack usage too close to ...". These operations seems to be necessary in the other geometries. 
```{r, echo=FALSE}
# 
# od_zones = st_read(file.path(folder, "Zonas_OD2017/Zonas_2017_region.shp"),  
#                    options = "ENCODING=LATIN1") %>%     
#   filter(NumeroMuni == 36) %>%
# #  st_buffer(dist = 0) %>%
#   st_transform(4326) %>%
#   st_make_valid()
od_zones = zonas_od

# Calcadas_SP = st_read(file.path(folder,"Poligonos_calcadas/Calcadas_SP.shp"))
# sw = Calcadas_SP %>%  # sidewalks
#   st_buffer(dist = 0) %>%
#   st_transform(4326) %>%
#   st_make_valid()
# saveRDS(sw, "sw.Rds")
# sf::write_sf(sw, "sw.gpkg")
sw = readRDS("sw.Rds")

# streets = st_read(file.path(folder,"Poligonos_vias/Poligonos_vias_total.gpkg")) %>%  # streets
#   st_buffer(dist = 0) %>%
#   st_transform(4326) %>%
#   st_make_valid()
# 
# area_sw = sw %>%
#   st_intersection(od_zones) %>%
#   mutate(area_sw = units::set_units(st_area(.), km^2)
#          ) %>%
#   select(NumeroZona, area_sw) %>%
#   st_drop_geometry() %>%
#   group_by(NumeroZona) %>%
#   summarise_all(sum)
# 
# area_streets = streets %>%
#   st_intersection(od_zones) %>%
#   mutate(area_streets = units::set_units(st_area(.), km^2)
#          ) %>%
#   select(NumeroZona, area_streets) %>%
#   st_drop_geometry() %>%
#   group_by(NumeroZona) %>%
#   summarise_all(sum)

```

```{r, echo=FALSE}

od_zones = od_zones %>%
  left_join(area_streets, by = c("NumeroZona")) %>%
  left_join(area_sw, by = c("NumeroZona")) %>%
  mutate(area = st_area(.),
         share_sw = as.numeric(area_sw/area)*100,
         share_streets = as.numeric(area_streets/area)*100
         )

```

```{r, echo=FALSE}

street = tm_shape(od_zones) +
  tm_fill(col = "area_streets",
          title = "Streets area (km^2)")

street_share = tm_shape(od_zones) +
  tm_fill(col = "share_streets",
          title = "Street share (%)")

sw = tm_shape(od_zones) +
  tm_fill(col = "area_sw",
          title = "Sidewalk area (km^2)")

sw_share = tm_shape(od_zones) +
  tm_fill(col = "share_sw",
          title = "Sidwalk share (%)")

tmap_arrange(street, sw, street_share, sw_share, ncol = 2)

```

OD pairs that generate the most intensive usage of space (kms traveled in the routes) have more area allocated to the transport mode (both foot and car) 

Very strong pattern here.

```{r, echo=FALSE}

foot_routes %>%
  filter(
    (zona_o %in% unique(OD_sp$zona_o)) & (zona_d %in% unique(OD_sp$zona_d))
    ) %>%
  left_join(area_sw, by = c("zona_o" = "NumeroZona")) %>%
  rename(sw_o = area_sw) %>%
  left_join(area_sw, by = c("zona_d" = "NumeroZona")) %>%
  rename(sw_d = area_sw) %>%
  ggplot(aes(x = log(as.numeric(0.5*sw_o+0.5*sw_d)),
             y = log(as.numeric(tot_length))
             )
         ) +
  geom_point() +
  xlab("log(OD mean sidewalk area)") +
  ylab("log(total km walked OD pair)") +
  theme_bw()

```

Some outliers in this graph, may be worth to check later...

```{r, echo=FALSE}

car_routes %>%
  filter(
    (zona_o %in% unique(OD_sp$zona_o)) & (zona_d %in% unique(OD_sp$zona_d))
    ) %>%
  left_join(area_streets, by = c("zona_o" = "NumeroZona")) %>%
  rename(streets_o = area_streets) %>%
  left_join(area_streets, by = c("zona_d" = "NumeroZona")) %>%
  rename(streets_d = area_streets) %>%
  ggplot(aes(x = log(as.numeric(0.5*streets_o+0.5*streets_d)),
             y = log(as.numeric(tot_length))
             )
         ) +
  geom_point() +
  xlab("log(OD mean streets area)") +
  ylab("log(total km drived in OD pair)") +
  theme_bw()

```
