---
title: "Exploring OD2017 data"
output: github_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)

library(tidyverse)
library(sf)
library(tmap)
library(stplanr)

load("C:/Users/Lucas/Downloads/OD2017/OD2017.rda")

```

I create this separated document to be the main reference on the exploratory data analysis. The data set is at the trip level and I take advantage of the survey design and use the expansion factor (for the trips) in all the graphs and analysis here. The survey implies that there are `r sum(OD_2017_v1$fe_via, na.rm=TRUE)` trips/day in the São Paulo Metro Region.

### Key variables

The data set contains a lot of socioeconomic variables. Since our interest relies on trips, flows and mode choices, the analyses so far are at the trip level. In this context, the variables that contain a the modes used in each trip are the main ones (`modo1`, `modo2`, `modo3`, and `modo4`). There is also the coordinates for each residence (`co_dom_x` and `co_dom_y`) and workplaces (`co_tr1_x` and `co_tr1_y`), and obviously the origin and destination zones and coordinates (e.g. `zona_o`, `co_o_x`, `co_o_y`) and euclidean distance. As you can see in the graph below, the vast majority of trips are made using just one mode. I use the main mode used in the trip (`modoprin`) in the remaining graphs. 

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  mutate(
    n_modes = case_when(
      !is.na(modo1) & is.na(modo2) & is.na(modo3) & is.na(modo4)    ~ "1",
      !is.na(modo1) & !is.na(modo2) & is.na(modo3) & is.na(modo4)   ~ "2",
      !is.na(modo1) & !is.na(modo2) & !is.na(modo3) & is.na(modo4)  ~ "3",
      !is.na(modo1) & !is.na(modo2) & !is.na(modo3) & !is.na(modo4) ~ "4"
    )
  )
  

OD_2017_v1 %>%
  ggplot() +
  geom_bar(aes(n_modes, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Number of modes used in each trip") +
  xlab("") +
  theme_bw()

```


### Different ways of defining modes

I aggregate the different names used for similar things. And this is the resulting distribution of trips for each mode.

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    modo_simples = case_when(
      modoprin==1 | modoprin==2 | modoprin==3 ~ "Metrô, trem e monotrilho",
      modoprin==4 | modoprin==5 | modoprin==6 ~ "Ônibus",
      modoprin==7 | modoprin==8            ~ "Fretado e Escolar",
      modoprin==9 | modoprin==10           ~ "Automóvel",
      modoprin==11 | modoprin==12          ~ "Táxi e derivados",
      modoprin==13 | modoprin==14          ~ "Moto",
      modoprin==15                      ~ "Bicicleta",
      modoprin==16                      ~ "A pé",
      modoprin==17                      ~ "Outros",
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(modo_simples, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Aggregating similar modes") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```


### Recategorisation of modes

I tried two different recategorisations (see code for the choices).The first is based on the "vehicle" itself, and the other relies on the "role" of the individual while in the trip.

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    mode_recat1 = case_when(
      modoprin %in% 1:6 ~ "Public transport",
      modoprin %in% 9:10 ~ "Car",
      modoprin %in% 13:14 ~ "Motorcycle",
      modoprin == 15 ~ "Bicycle",
      modoprin == 16 ~ "On foot",
      modoprin == 17 | modoprin == 12 | modoprin == 11 | modoprin == 7 | modoprin == 8 ~ "Other"
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(mode_recat1, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Recategorisation of modes 1") +
  xlab("")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```

```{r, echo=FALSE}

OD_2017_v1 = OD_2017_v1 %>%
  dplyr::mutate(
    mode_recat2 = case_when(
      modoprin %in% 1:6 ~ "Public transport",
      modoprin == 9 | modoprin == 13 ~ "Driving",
      modoprin==10 | modoprin==14 ~ "Rider",
      modoprin %in% c(7, 8, 11, 12) ~ "Private transport passenger",
      modoprin==15 ~ "Cycling",
      modoprin==16 ~ "Walking",
      modoprin==17 ~ "Other"
    )
  )

OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  ggplot() +
  geom_bar(aes(mode_recat2, weight=fe_via)) +
  scale_y_continuous(name = "Thousands of trips / day",
                     labels=function(x) format(x/1000,
                                               big.mark = ",",
                                               decimal.mark = ".",
                                               scientific = FALSE)
                     ) +
  ggtitle("Recategorisation of modes - 2") +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1))

```


### Basic geographic analysis

```{r, include=FALSE}

zonas_od = st_read("C:/Users/Lucas/Downloads/OD2017/shapefiles/Zonas_2017_region.shp")
tmap_mode("plot")

```

Here I plot the main zones of origin and destination using the total number of trips -- always using the survey's expansion factor. It is interesting to note that there is no centrality pattern when we look at the main Zones of Destination. Maybe we should restrict the analysis to specific types of trips (e.g., commute strictly defined). I checked these results with the official Tables and they are the same in terms of the main zones of origin and destination.

```{r, echo=FALSE}

OD = OD_2017_v1 %>%
  filter(!is.na(modoprin)) %>%
  select(zona_o, zona_d, modoprin, fe_via) %>%
  mutate(
    public = ifelse(modoprin %in% 1:6, 1, 0)*fe_via,
    car = ifelse(modoprin %in% 9:10, 1, 0)*fe_via,
    motorcycle = ifelse(modoprin %in% 13:14, 1, 0)*fe_via,
    bike = ifelse(modoprin == 15, 1, 0)*fe_via,
    foot = ifelse(modoprin == 16, 1, 0)*fe_via,
    other = ifelse(modoprin %in% c(7, 8, 11, 12, 17), 1, 0)*fe_via
         ) %>%
  select(zona_o, zona_d, public, car, motorcycle, bike, foot, other) %>%
  group_by(zona_o, zona_d) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  mutate(trips = public + car + motorcycle + bike + foot + other)

# check:
# sum(OD$public) + sum(OD$car) + sum(OD$motorcycle) + sum(OD$bike) + sum(OD$foot) + sum(OD$other)

origin = OD %>%
  group_by(zona_o) %>%
  summarise_all(sum) %>%
  select(zona_o, trips_o = trips)

destination = OD %>%
  group_by(zona_d) %>%
  summarise_all(sum) %>%
  select(zona_d, trips_d = trips)

zonas_od_all = zonas_od %>%
  left_join(origin, by = c("NumeroZona" = "zona_o")) %>%
  left_join(destination, by = c("NumeroZona" = "zona_d"))


tm_shape(zonas_od_all) + tm_fill(c("trips_o", "trips_d"), 
                             style = "cont",
                             title = "Trips")  +
  tm_facets(free.scales = FALSE, ncol = 1, nrow = 2) +
  tm_layout(title = "São Paulo Metro Region - All trips",
            panel.labels = c("Zone of origin", "Zone of destination"),
            inner.margins = c(0, .02, .02, .02),
            legend.outside = TRUE)

```

Plotting only the São Paulo City.There is no pattern of spatial centrality in destinations either.

```{r, echo=FALSE}

zonas_od_sp_all = zonas_od_all %>%
  filter(NomeMunici=="São Paulo")

tm_shape(zonas_od_sp_all) + tm_fill(c("trips_o", "trips_d"),
                                style = "cont",
                                title = "Trips"
                                )  +
  tm_facets(free.scales = FALSE, ncol = 2) +
  tm_layout(title = "São Paulo City - All trips",
            panel.labels = c("Zone of origin", "Zone of destination"),
            inner.margins = c(0, .02, .02, .02))


```

I then remove the intrazonal trips and plot the same maps to test whether these patterns come from intrazonal trips only. It seems that is not the case.

```{r, echo=FALSE}

origin_inter = OD %>%
  filter(zona_o != zona_d) %>%
  group_by(zona_o) %>%
  summarise_all(sum) %>%
  select(zona_o, trips_o = trips)

destination_inter = OD %>%
  filter(zona_o != zona_d) %>%
  group_by(zona_d) %>%
  summarise_all(sum) %>%
  select(zona_d, trips_d = trips)

zonas_od_inter = zonas_od %>%
  left_join(origin_inter, by = c("NumeroZona" = "zona_o")) %>%
  left_join(destination_inter, by=c("NumeroZona" = "zona_d"))


tm_shape(zonas_od_inter) + tm_fill(c("trips_o", "trips_d"), 
                             style = "cont",
                             title = "Trips")  +
  tm_facets(free.scales = FALSE, ncol = 1, nrow = 2) +
  tm_layout(title = "São Paulo Metro Region - Interzones trips",
            panel.labels = c("Zone of origin", "Zone of destination"),
            legend.outside = TRUE,
            inner.margins = c(0, .02, .02, .02))


```

And the same conclusion applies for São Paulo City...

```{r, echo=FALSE}

zonas_od_sp_inter = zonas_od_inter %>%
  filter(NomeMunici=="São Paulo")

tm_shape(zonas_od_sp_inter) + tm_fill(c("trips_o", "trips_d"),
                                style = "cont",
                                title = "Trips"
                                )  +
  tm_facets(free.scales = FALSE, ncol = 2) +
  tm_layout(title = "São Paulo City - Interzones trips",
            panel.labels = c("Zone of origin", "Zone of destination"),
            inner.margins = c(0, .02, .02, .02))


```


#### Active travels and trip patterns

Following the Bristol examples in Robin's book now. But maybe we should zoom in the São Paulo City.

```{r, echo=FALSE}

od_inter = OD %>%
  filter(zona_o != zona_d) %>%
  mutate(
    active = (bike + foot)/trips * 100
  )

desire_lines = od2line(od_inter, zonas_od)

tm_shape(desire_lines) +
  tm_lines(palette = "plasma", breaks = c(0, 5, 10, 20, 40, 100),
    lwd = "trips",
    scale = 9,
    title.lwd = "Number of trips",
    alpha = 0.6,
    col = "active",
    title = "Active travel (%)"
  ) +
  tm_scale_bar()

```

There are still too many zones...

```{r, echo=FALSE}

od_inter_sp = OD %>%
  left_join(zonas_od, by=c("zona_o" = "NumeroZona")) %>%
  rename(muni_o = NomeMunici) %>%
  left_join(zonas_od, by=c("zona_d" = "NumeroZona")) %>%
  rename(muni_d = NomeMunici) %>%
  filter(zona_o != zona_d & muni_o == "São Paulo" & muni_d == "São Paulo") %>%
  mutate(
    active = (bike + foot)/trips * 100
  )

zonas_od_sp = zonas_od %>%
  filter(NomeMunici == "São Paulo")

desire_lines_sp = od2line(od_inter_sp, zonas_od_sp)

tm_shape(desire_lines_sp) +
  tm_lines(palette = "plasma", breaks = c(0, 5, 10, 20, 40, 100),
    lwd = "trips",
    scale = 9,
    title.lwd = "Number of trips",
    alpha = 0.6,
    col = "active",
    title = "Active travel (%)"
  ) +
  tm_scale_bar()

```
